New lower bounds for Schur and weak Schur numbers

### 2021-04-12 article/article.tex
rafinements + prop WS^+

This paper provides new lower bounds for both Schur and weak Schur numbers. These results were produced by pushing forward
Rowley's "templates"-based approach for Ramsey and Schur numbers in 2020. Finding suitable templates allows us to apply Rowley's construction
to get explicit partitions improving lower bounds as well as the growth rate of Schur numbers and Ramsey numbers \(R_n(3)\). 
We also developed a new method to improve lower bounds on weak Schur numbers. Furthermore, this paper tries to analyze former 
works on the subject based on the principle that good partitions into \(n+1\) subsets start with good partitions 
into \(n\) subsets. We show that exceeding the previous lower bound \(WS(6) \geqslant 582\) is impossible with such an assumption upon 
imposing certain conditions on the good 5-subsets partition. The new lower bounds includes \(S(9) \geqslant 17803\), \(S(10) \geqslant 60948\), 
\(WS(9) \geqslant 22536 \) and \(WS(10) \geqslant 71214 \).

>>>>

This paper provides new lower bounds for both Schur and weak Schur numbers. These results were produced by pushing forward
Rowley's "templates"-based approach for Ramsey and Schur numbers in 2020. Finding suitable templates allows us to apply Rowley's construction
to get explicit partitions improving lower bounds as well as the growth rate of both Schur numbers and Ramsey numbers \(R_n(3)\). 
We also developed a new method to improve lower bounds on weak Schur numbers. Furthermore, this paper tries to analyze former 
works on the subject based on the principle that good partitions into \(n+1\) subsets start with good partitions 
into \(n\) subsets. We show that exceeding the previous lower bound \(WS(6) \geqslant 582\) is impossible with such an assumption upon 
imposing certain conditions on the good 5-subsets partition. The new lower bounds includes \(S(9) \geqslant 17803\), \(S(10) \geqslant 60948\), 
\(WS(9) \geqslant 22536 \) and \(WS(10) \geqslant 71214 \).

### 2021-04-12 article/schur.tex
rafinements + prop WS^+

The idea lying beneath this inequality is similar to Abbott and Hanson's contruction \cite{AbbottHanson}. 
They extend vertically a sum-free partition, and horizontally an other sum-free partition. This way each "block" 
acts like a security zone for the other one. Here, the horizontal partition is no longer to the side of the 
vertical one, but it occupies the column of the special color of the \(SF\)-template, i.e the one without the 
extra condition. We give the following example for \(p = 7, q = 4, n = 2 ~\textrm{and}~ k = 2\). \\

@@

\textsc{Proof :} We denote by \(f\) the colouring associated to the \(SF\)-template of lenght \(p\)
and \(g\) the one associated to the sum-free partition of \([\![1,q]\!]\).

@@

Using a SAT solver, we exhibited SF-templates, hence providing lower bound on \(S^+\) and inequalities 
of the type \(S(n+k) \geqslant a S(n) + b\). We have sought templates providing the greatest value of 
\((a, b)\) (for the lexicographic order). When the number of colors excceeded 5,in order to reduce the search space we 
looked for symmetric SF-templates, we assumed that the special color was the last color to appear and we constrained 
the \(m_c\)'s out of being too small. Further details about the encoding as a SAT problem can be found in the
\hyperref[SAT]{SAT section}. \\

@@

The first 3 inequalities are optimal. The fourth one is optimal among symmetric SF-templates whose special color is 
the last in the order of apparition (and with a multiplicative factor less than or equal to 118). The fifth one is most 
likely not optimal but should not be too far from the optimal SF-template. 
Finally, the sixth one is obtained by combining (see below) the SF-template of length 380 and the one of length 3. Although we 
could not find a better SF-template with 7 colors, this one definetely very far from the optimal value. One may try to seek better 
templates by constraining less the search space and by using a Monte-Carlo method, as in \cite{Bouzy2015AnAP}. 
This could be the suject of a future work.\\

>>>>

\begin{remark}
The following proposition can help improve the additive constant of a SF-template. Although it does not allow us to improve the SF-templates 
we have found, the analogous of this proposition for WSF-templates (see next section) allows to improve one of them.
\end{remark}

\begin{proposition}
Let \((k, p) \in \mathbb{N}^*)^2\) and let \(f\) be a colouring associated to a SF-template of length \(p\) with \(k\) colors. Let 
\(b \in \mathbb{N}\) (\(b = m_{k+1} - 1\) works) and assume there there exists a colouring \(g\) of \([\![1, b]\!]\) with \(k\) colors such that:
\begin{itemize}
	\item \(\forall c \in [\![1, k]\!], \forall (x, y) \in [\![1, p]\!]^2,(f(x) = f(y) \text{ and } (x + y) \mod p \leqslant b) 
	\implies g((x + y) \mod p) \neq f(x)\)
	\item \(\forall c \in [\![1, k]\!], \forall (x, y) \in [\![1, p]\!] \times  [\![1, b]\!], (f(x) = g(y) \text{ and } x + y \leqslant b) 
	\implies g(x + y) \neq f(x)\)
\end{itemize}
Then, for every \(n \in \mathbb{N}^*\), by using on the last row the colouring \(i \longmapsto g(i - p S(n))\), we have\\
\[ S(n+k) \geqslant S^+(k+1)S(n) + b\]
\end{proposition}

\begin{remark}
This proposition corresponds to the fact that sometimes a column is not the sum of two columns of a given color, but adding this 
column to the color would create sums in the color when applying the extension procedure. However, the last line does not 
intereact will all the columns when it comes to creating new sums. As a result, the hypotheses made on the colouring of the last 
row can be weakened.
\end{remark}

The idea lying beneath this theorem is similar to Abbott and Hanson's contruction \cite{AbbottHanson}. 
They extend vertically a sum-free partition, and horizontally an other sum-free partition. This way each "block" 
acts like a security zone for the other one. Here, the horizontal partition is no longer to the side of the 
vertical one, but it occupies the column of the special color of the \(SF\)-template, i.e the one without the 
extra condition. We give the following example for \(p = 7, q = 4, n = 2 ~\textrm{and}~ k = 2\). \\

@@

\textsc{Proof of theorem:} We denote by \(f\) the colouring associated to the \(SF\)-template of lenght \(p\)
and \(g\) the one associated to the sum-free partition of \([\![1,q]\!]\).

@@

Using a SAT solver, we exhibited SF-templates, hence providing lower bound on \(S^+\) and inequalities 
of the type \(S(n+k) \geqslant a S(n) + b\). We have sought templates providing the greatest value of 
\((a, b)\) (for the lexicographic order). When the number of colors excceeded 5, in order to reduce the search space we 
looked for symmetric SF-templates, we assumed that the special color was the last color to appear and we constrained 
the \(m_c\)'s out of being too small. Further details about the encoding as a SAT problem can be found in the
\hyperref[SAT]{SAT section}. \\

@@

The first 3 inequalities are optimal. The fourth one is optimal among symmetric SF-templates whose special color is 
the last in the order of apparition (and with a multiplicative factor less than or equal to 118). The fifth one is most 
likely not optimal but should not be too far from the optimal SF-template. 
Finally, the sixth one is obtained by combining (see below) the SF-template of length 380 and the one of length 3. 
Although we could not find a better SF-template with 7 colors, the last inequality is definetely very far from the 
optimal value. One may try to seek better templates by constraining less the search space and by using 
Monte-Carlo methods, as in \cite{Bouzy2015AnAP}. This could be the suject of a future work.\\

### 2021-04-12 article/weakschur.tex
petit oubli
rafinements + prop WS^+

\begin{itemize}
	\item \(\mathcal{T}\) (the "tail"): the integers from 1 to \(q\). NB: this is line number 0.
	\item \(\mathcal{L}\) (the "lines"): the integers in columns \(-l\) to \(+r\) (excluding  \(\mathcal{T}\)).
	\item \(\mathcal{C}\) (the "columns"): the integers in the last \(w\) columns (excluding  \(\mathcal{T}\)).
\end{itemize}

Like for SF-templates, \(\mathcal{L}\) and \(\mathcal{C}\) play the role of security zones for each other. Note that with 
this numbering of columns, the column of the sum of two numbers is the only integer in \([\![-l,q]\!]\) equal to two the 
sum of the columns modulo \(a\).

@@

\underline{\textbf{\(\mathcal{L}\) block}} \\
In this block,  we use the colors \(n + 1, ..., n + k\). We colour an integer \(x\) according to its line number (written \(\lambda(x)\)).
For every \(x \in \mathcal{L}\), we colour \(x\) with \(n + c\) where \(c\) is the colour of \(\lambda(x)\) in the sum-free colouring of  \([\![1,p]\!]\).
Let \((x, y) \in \mathcal(L)^2\). The cases are twofold.

@@

\begin{remark}
This formula includes the results of Rowley \cite{RowleyWS} as a special case. For \(n>2\),
this formula does not give new lower bounds but in the same way as we introduced \(S^+\)\hyperref[SE]{(Definition
3.1)},we will define \(WS^+\) and find inequalities between \(WS^+\),\(WS\) and \(S\)
\end{remark}

@@

We will now define functions and results, that will be used in the following theorem.

@@

@@

\begin{corollary}
\( \forall (n,k) \in (\mathbb{N}^*)^2 \text{, } let ~ b_{max}=max \{b\in \mathbb{N}^*/WS^+(n+1)=WS_b^+(n+1)\},\) \[
WS(n+k) \geqslant S(k) WS^+(n+1)+b_{max}\]
\end{corollary}

@@

\begin{remark}
There exists a \(c(b) \geqslant \text{min}(A_{k+1} \backslash [\![1,b] - b - 1\) such that \(c(b)\) numbers can be
added at the end of the extended partition.Therefore, we can get a better lower bound by finding a couple \((b,c(b))\)
which maximizes the sum \(b+c(b)\) such that \(WS^+(k+1)=WS_b^+(k+1)\). Hence we have:
	\[ WS(n+k) \geqslant S(n) WS^+(k+1)+b+c(b)\]

\end{remark}

@@

@@

\[
4S(n) + 1 \leqslant WS(n+1)
\]
\[
13S(n) + 8 \leqslant WS(n+2)
\]
\[
42S(n) + 24 \leqslant WS(n+3)
\]
\[
127S(n) + 68 \leqslant WS(n+4)
\]

The first two inequalities were found by Rowley, they are detailed in [2]. The third inequality is optimal and was found with a SAT solver. 
As for the fourth inequality, it is not useful since by comibing the first inequality with \(S(n) \geqslant 33 S(n-3) + 6\) provides a 
better inequality. It was also found with the SAT solver. In order to reduce the search space, we only looked for WSF-templates of 
5 colors which start with a good \(WS(4)\) partition. However, this approach most likely prevents us from finding the best WSF-templates 
as we explain in the next subsection for weakly sum-free partitions. We highly suspect that there exists more efficient WSF-templates 
with \(n \geqslant 5\) colors. One may try to go over a different search space using a Monte-Carlo method, as in \cite{Bouzy2015AnAP}. 
This could be the suject of a future work.

>>>>

\begin{itemize}
	\item \(\mathcal{T}\) (the "tail"): the integers from 1 to \(q\). NB: this is line number 0.
	\item \(\mathcal{R}\) (the "rows"): the integers in columns \(-l\) to \(+r\) (excluding  \(\mathcal{T}\)).
	\item \(\mathcal{C}\) (the "columns"): the integers in the last \(w\) columns (excluding  \(\mathcal{T}\)).
\end{itemize}

Like for SF-templates, \(\mathcal{R}\) and \(\mathcal{C}\) play the role of security zones for each other. Note that with 
this numbering of columns, the column of the sum of two numbers is the only integer in \([\![-l,q]\!]\) equal to two the 
sum of the columns modulo \(a\).

@@

\underline{\textbf{\(\mathcal{R}\) block}} \\
In this block,  we use the colors \(n + 1, ..., n + k\). We colour an integer \(x\) according to its line number (written \(\lambda(x)\)).
For every \(x \in \mathcal{R}\), we colour \(x\) with \(n + c\) where \(c\) is the colour of \(\lambda(x)\) in the sum-free colouring of  \([\![1,p]\!]\).
Let \((x, y) \in \mathcal{R}^2\). The cases are twofold.

@@

\begin{remark}
This formula includes the results of Rowley \cite{RowleyWS} as a special case. For \(n>2\),
this formula does not give new lower bounds but in the same way as we introduced \(S^+\)\hyperref[SE]{(Definition
3.1)},we define \(WS^+\) and find inequalities between \(WS^+\),\(WS\) and \(S\)
\end{remark}

@@

In this subsection, we define the objects and proove the results needed for the general theorem regarding templates for 
weak Schur numbers.

@@

\begin{proposition}
	Let \(n \in [\![2, +\infty]\!]\), we have :
	\[
	\frac{3}{2}WS(n-1)+1 \leqslant WS^+(n) \leqslant WS(n)
	\]
\end{proposition}

\textsc{Proof :} The lower bound comes from the analogous of Abott and Hanson's construction for weak Schur numbers. 
The upper bound comes from the fact that a WSF-template of length \(p\) with \(n\) colors is also a partition of
\([\![1, p]\!]\) into \(n\) sum-free subsets.

\begin{remark}
	\(WS^+\) and \(WS\) have the same asymptotic growth rate.
\end{remark}

@@

\begin{corollary}
\( \forall (n,k) \in (\mathbb{N}^*)^2 \text{, } let ~ b_{max}=max \{b\in \mathbb{N}^*/ WS_b^+(n+1) = WS^+(n+1)\},\) \[
WS(n+k) \geqslant S(k) WS^+(n+1)+b_{max}\]
\end{corollary}

\begin{remark}
In the SF-template construction for Schur numbers, the additive constant comes from the fact that the special color does 
not necessarily appear right at the begining of the repeating pattern. Likewise, \(b_max\) can actually be replaced by \\
\[\max \left\{ b + \min (A_{n+1} \backslash [\![1, b ]\!]) - 1~|~ WS_b^+(n+1) = WS^+(n+1) \right\}\]
\end{remark}

\begin{remark}
Acutally, like for SF-templates, the additive constant of a WSF-template (in the form given by the above remark) can be improved 
by weakening the hypotheses made on the last row. The principle behind it is the same as in the analogous proposition for SF-templates.
\end{remark}

\begin{proposition}
Let \((b, k, p) \in \mathbb{N}^*)^3\) and let \(f\) be a colouring associated to a b-WSF-template of length \(p\) with \(k\) colors. Let 
\(c \in \mathbb{N}\) (\(c = \min (A_{k+1} \backslash [\![1, b ]\!]) - 1\) works) and assume there there exists a colouring \(g\) of 
\([\![b + 1, b + c]\!]\) with \(k\) colors such that:
\begin{itemize}
	\item \(\forall c \in [\![1, k]\!], \forall (x, y) \in  [\![1, a + b]\!] \times  [\![b + 1, a + b]\!], (f(x) = f(y) \text{ and } \pi(x + y) \leqslant b + c) 
	\implies g(\pi(x + y)) \neq f(x)\)
	\item \(\forall c \in [\![1, k]\!], \forall (x, y) \in  [\![1, a + b]\!] \times  [\![b + 1, b + c]\!],  (f(x) = g(y) \text{ and } \pi(x + y) \leqslant b + c) 
\implies g(\pi(x + y)) \neq f(x)\)
\end{itemize}
Then, for every \(n \in \mathbb{N}^*\), by using on the last row the colouring \(i \longmapsto g(i - p S(n)\), we have\\
\[ WS(n+k) \geqslant WS^+(k+1) S(n) + b + c\]
\end{proposition}

@@

@@

\begin{remark}
The WSF-templates can actually be fine-tuned further. However, it gives only minor improvements (most likely only additive 
constant) at the cost of dramatically increasing the size of the search space while. Therefore, it does not seem relevant to 
use this sophistications given that we could not even find good WSF-templates with 5 colors using a computer (here good 
means better than those obtain by combining smaller templates).

\par These modifications work as follows. One may notice that the first row (excluding the "tail") has constraints that other rows 
do not have because of the tail, especially if the special color appears in the tail as well. Thus allowing to have a colouring on the 
first row different from the colouring of the other rows would weaken the constraints. Acutally, one may even go further by 
noticing that on the one hand the first (ordered) color of the sum-free partition used for the extension procedure has more 
more constraints than the other colors of the sum-free partition since the first row is of this color and is more constrained than 
the other rows, but that on the other hand it has more degrees of freedom than the other colors of the sum-free partition since in 
the sum-free partition there cannot be two consecutive numbers of this color. As a result, it removes  some constraints imposed
 by the first row on the other rows.

\par To sum up, one can look for a generalised WSF-template that uses a special colouring for the tail and the first row, a 
colouring dedicated to the rows whose number is not 1 but is in the first color in the sum-free partition, a colouring for all 
the other rows and a special colouring for the last numbers (as previously explained for the improvement of the additive 
constant of WSF-templates).
\end{remark}

We also have a similar theorem where only \(WS^+\) is involved.

\begin{theorem}
	Let \((n,k), (p,q) \in (\mathbb{N}^*)^2\). If there exists a \(SF\)-template of \(k+1\) colors and lenght \(p\),
	and \(WSF\)-template of \(n\) color and lenght \(q\), then there exists \(WSF\)-template of \((n+k)\) and lenght \(pq\).
\end{theorem}

And the associated inequality :

\begin{corollary}
	Let \(n, k \in \mathbb{N}^*\), we have \\
	\[ WS^+(n+k) \geqslant S^+(k+1)WS^+(n) \]
\end{corollary}

\textsc{Proof :} The idea is the same as in the previous theorem. The only difference is the \(SF\) property inherited 
from the second \(SF\)-template.

@@

\[
4S(n) + 1 \leqslant WS(n+1)
\]
\[
13S(n) + 8 \leqslant WS(n+2)
\]
\[
42S(n) + 24 \leqslant WS(n+3)
\]
\[
132S(n) + 26 \leqslant WS(n+4)
\]

The first two inequalities were found by Rowley, they are detailed in [2]. The third inequality is optimal and was found with a SAT solver. 
It uses the first sophistication explained in the previous subsection in order to add the last number in the first color.  
As for the fourth inequality, it was obtained by combining an optimal SF-template of length 33 with a WSF-template of length 4. 
The best template we could get with a computer search gives the inequality \(WS(n+4) \geqslant 127 S(n) + 68\). 
It was also found with the SAT solver. In order to reduce the search space, we only looked for WSF-templates of 
5 colors which start with a good \(WS(4)\) partition. However, this approach most likely prevents us from finding the best WSF-templates 
as we explain in the next subsection for weakly sum-free partitions. We highly suspect that there exists more efficient WSF-templates 
with \(n \geqslant 5\) colors. One may try to go over a different search space using a Monte-Carlo method, as in \cite{Bouzy2015AnAP}. 
This could be the suject of a future work. Further details about the encoding as a SAT problem can be found in the
\hyperref[SAT]{SAT section}. \\
